Faithful Model Explanations through\\ Energy-Constrained Conformal Counterfactuals\footnote{This is a pre-print.}

### 2024-01-10 paper/camera-ready/altmeyer.tex
tables sorted

For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}~\citep{lecun1998gradient}). Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}. To account for stochasticity, we generate many counterfactuals for each target class, generator, model and dataset over multiple runs.

@@

Full details concerning model training as well as detailed descriptions and results for all datasets can be found in the appendix. In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}. The tables show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}). For each metric, the best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively. The third metric presented in Table~\ref{tab:results-tabular} quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}. For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.

@@

\begin{table*}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llcccccc}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{3}{c}{California Housing} & \multicolumn{3}{c}{GMSC} \\
\cmidrule(l{3pt}r{3pt}){3-5} \cmidrule(l{3pt}r{3pt}){6-8}
Model & Generator & Unfaithfulness ↓ & Implausibility ↓ & Uncertainty ↓ & Unfaithfulness ↓ & Implausibility ↓ & Uncertainty ↓\\
\midrule
    & ECCCo & \textbf{3.69 ± 0.08}** & 1.94 ± 0.13\hphantom{*}\hphantom{*} & \textbf{0.09 ± 0.01}** & 3.84 ± 0.07** & 2.13 ± 0.08\hphantom{*}\hphantom{*} & \textbf{0.23 ± 0.01}**\\

@@

\multirow{-7}{*}{\raggedright\arraybackslash MLP Ensemble} & Wachter & 4.04 ± 0.07\hphantom{*}\hphantom{*} & 1.13 ± 0.12\hphantom{*}\hphantom{*} & 0.16 ± 0.01\hphantom{*}\hphantom{*} & 4.10 ± 0.07\hphantom{*}\hphantom{*} & 0.95 ± 0.08\hphantom{*}\hphantom{*} & 0.32 ± 0.01\hphantom{*}\hphantom{*}\\
\cmidrule{1-8}
    & ECCCo & 1.40 ± 0.08** & 0.69 ± 0.05** & 0.11 ± 0.00** & 1.20 ± 0.06*\hphantom{*} & 0.78 ± 0.07** & 0.38 ± 0.01\hphantom{*}\hphantom{*}\\

@@

\multirow{-7}{*}{\raggedright\arraybackslash JEM Ensemble} & Wachter & 1.71 ± 0.09\hphantom{*}\hphantom{*} & 0.99 ± 0.08\hphantom{*}\hphantom{*} & 0.14 ± 0.00\hphantom{*}\hphantom{*} & 1.31 ± 0.08\hphantom{*}\hphantom{*} & 0.95 ± 0.10\hphantom{*}\hphantom{*} & 0.33 ± 0.01\hphantom{*}\hphantom{*}\\
\bottomrule
\end{tabular}}
\caption{Results for tabular datasets: sample averages +/- one standard deviation across valid counterfactuals. The best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\textit{Wachter}). \label{tab:results-tabular}}
\end{table*}

@@

We also generally find that latent space search does not impede faithfulness for \textit{ECCCo}. In most cases \textit{ECCCo+} is either on par with \textit{ECCCo} or even outperforms it. There are some notable exceptions though. Cases in which \textit{ECCCo} achieves substantially better faithfulness without latent space search tend to involve more vulnerable models like the simple MLP for MNIST in Table~\ref{tab:results-vision}. We explain this finding as follows: even though dimensionality reduction through PCA in the case of \textit{ECCCo+} can be considered a relatively mild form of intervention, the first $n_z$ principal components fail to capture some of the variation in the data. More vulnerable models may be particularly sensitive to this residual variation in the data.

@@

\begin{table}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llcc}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{MNIST} \\
\cmidrule(l{3pt}r{3pt}){3-4}
Model & Generator & Unfaithfulness ↓ & Implausibility ↓\\
\midrule
    & ECCCo & \textbf{0.243 ± 0.000}** & 0.420 ± 0.001\hphantom{*}\hphantom{*}\\

@@

\multirow{-5}{*}{\raggedright\arraybackslash LeNet-5} & Wachter & 0.249 ± 0.000\hphantom{*}\hphantom{*} & 0.335 ± 0.002\hphantom{*}\hphantom{*}\\
\bottomrule
\end{tabular}}
\caption{Results for vision dataset. Formatting details are the same as in Table~\ref{tab:results-tabular}. \label{tab:results-vision}}
\end{table}

@@

Firstly, we recognize that our proposed distance-based evaluation metrics for plausibility and faithfulness may not be universally applicable to all types of data. In any case, they depend on choosing a distance metric on a case-by-case basis, as we have done in this work. Arguably, commonly used metrics for measuring other desiderata such as closeness suffer from the same pitfall. We therefore think that future work on counterfactual explanations could benefit from defining universal evaluation metrics.

>>>>

For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}~\citep{lecun1998gradient}). Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to enhance the plausibility of \textit{ECCCo}. To account for stochasticity, we generate many counterfactuals for each target class, generator, model and dataset over multiple runs.

@@

Full details concerning model training as well as detailed descriptions and results for all datasets can be found in the appendix. In the following, we focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}. They show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}). For each metric, the best outcomes are in bold. Asterisks indicate that a given value is more than one (*) or two (**) standard deviations away from the baseline (\textit{Wachter}). For the tabular datasets, we use the Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively. For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}. The third metric presented in Table~\ref{tab:results-tabular} quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}.

@@

\begin{table*}
  \setlength{\tabcolsep}{4pt}
  \centering
  \fontsize{9}{11}\selectfont
  \begin{tabular}[t]{llcccccc}
  \toprule
  \multicolumn{2}{c}{ } & \multicolumn{3}{c}{California Housing} & \multicolumn{3}{c}{GMSC} \\
  \cmidrule(l{3pt}r{3pt}){3-5} \cmidrule(l{3pt}r{3pt}){6-8}
  Model & Generator & Unfaithfulness ↓ & Implausibility ↓ & Uncertainty ↓ & Unfaithfulness ↓ & Implausibility ↓ & Uncertainty ↓\\
  \midrule
   & ECCCo & \textbf{3.69 ± 0.08}** & 1.94 ± 0.13\hphantom{*}\hphantom{*} & \textbf{0.09 ± 0.01}** & 3.84 ± 0.07** & 2.13 ± 0.08\hphantom{*}\hphantom{*} & \textbf{0.23 ± 0.01}**\\

@@

  \multirow{-7}{*}{\raggedright\arraybackslash MLP} & Wachter & 4.04 ± 0.07\hphantom{*}\hphantom{*} & 1.13 ± 0.12\hphantom{*}\hphantom{*} & 0.16 ± 0.01\hphantom{*}\hphantom{*} & 4.10 ± 0.07\hphantom{*}\hphantom{*} & 0.95 ± 0.08\hphantom{*}\hphantom{*} & 0.32 ± 0.01\hphantom{*}\hphantom{*}\\
  \cmidrule{1-8}
   & ECCCo & 1.40 ± 0.08** & 0.69 ± 0.05** & 0.11 ± 0.00** & 1.20 ± 0.06*\hphantom{*} & 0.78 ± 0.07** & 0.38 ± 0.01\hphantom{*}\hphantom{*}\\

@@

  \multirow{-7}{*}{\raggedright\arraybackslash JEM} & Wachter & 1.71 ± 0.09\hphantom{*}\hphantom{*} & 0.99 ± 0.08\hphantom{*}\hphantom{*} & 0.14 ± 0.00\hphantom{*}\hphantom{*} & 1.31 ± 0.08\hphantom{*}\hphantom{*} & 0.95 ± 0.10\hphantom{*}\hphantom{*} & 0.33 ± 0.01\hphantom{*}\hphantom{*}\\
  \bottomrule
  \end{tabular}
\caption{Results for tabular datasets: sample averages +/- one standard deviation across valid counterfactuals. Model ensembling was used in both cases. The best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\textit{Wachter}). \label{tab:results-tabular}}
\end{table*}

@@

We also generally find that latent space search does not impede faithfulness for \textit{ECCCo}. In most cases \textit{ECCCo+} is either on par with \textit{ECCCo} or outperforms it. There are some notable exceptions though. Cases in which \textit{ECCCo} achieves substantially better faithfulness without latent space search tend to involve more vulnerable models like the MLP for MNIST in Table~\ref{tab:results-vision}. We explain this finding as follows: even though dimensionality reduction through PCA in the case of \textit{ECCCo+} can be considered a relatively mild form of intervention, the first $n_z$ principal components fail to capture some of the variation in the data. More vulnerable models may be particularly sensitive to this residual variation in the data.

@@

\begin{table}
  \centering
  \fontsize{9}{11}\selectfont
  \begin{tabular}[t]{llcc}
  \toprule
  \multicolumn{2}{c}{ } & \multicolumn{2}{c}{MNIST} \\
  \cmidrule(l{3pt}r{3pt}){3-4}
  Model & Generator & Unfaithfulness ↓ & Implausibility ↓\\
  \midrule
   & ECCCo & \textbf{0.243 ± 0.000}** & 0.420 ± 0.001\hphantom{*}\hphantom{*}\\

@@

  \multirow{-5}{*}{\raggedright\arraybackslash LeNet-5} & Wachter & 0.249 ± 0.000\hphantom{*}\hphantom{*} & 0.335 ± 0.002\hphantom{*}\hphantom{*}\\
  \bottomrule
  \end{tabular}
\caption{Results for vision dataset. No model ensembling was used. Formatting details are the same as in Table~\ref{tab:results-tabular}. \label{tab:results-vision}}
\end{table}

@@

Firstly, we recognize that our proposed distance-based evaluation metrics for plausibility and faithfulness may not be universally applicable to all types of data. They depend on choosing a distance metric on a case-by-case basis, as we have done in this work. Arguably, commonly used metrics for measuring other desiderata such as closeness suffer from the same pitfall. We therefore think that future work could benefit from defining universal evaluation metrics.

### 2024-01-10 paper/tables.Rmd
tables sorted

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i[,variable:=factor(variable, levels=measures)]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i <- tab_i[model %in% chosen_model]
tab_i[,model:=factor(model,levels=chosen_model)]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages +/- one standard deviation across valid counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\\textit{Wachter}). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep, table.env="table*"
) %>%
  kable_styling(
    latex_options = c("scale_down")
  ) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

@@

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i[,variable:=factor(variable, levels=measures)]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i <- tab_i[model %in% chosen_model]
tab_i[,model:=factor(model,levels=chosen_model)]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s dataset. Formatting details are the same as in Table~\\ref{tab:results-tabular}. \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(
    latex_options = c("scale_down")
  ) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

>>>>

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i[,variable:=factor(variable, levels=measures)]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i <- tab_i[model %in% chosen_model]
tab_i[,model:=factor(model,levels=chosen_model)]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)

# Change model names to reduce column space
levels(tab_i$model)[match("MLP Ensemble",levels(tab_i$model))] <- "MLP"
levels(tab_i$model)[match("JEM Ensemble",levels(tab_i$model))] <- "JEM"

col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages +/- one standard deviation across valid counterfactuals. Model ensembling was used in both cases. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\\textit{Wachter}). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep, table.env="table*"
) %>%
  kable_styling(
    font_size = 9
  ) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

@@

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i[,variable:=factor(variable, levels=measures)]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i <- tab_i[model %in% chosen_model]
tab_i[,model:=factor(model,levels=chosen_model)]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s dataset. No model ensembling was used. Formatting details are the same as in Table~\\ref{tab:results-tabular}. \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(
    font_size = 9
  ) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```
